Modeling the impacts of replacing car trips with combined public transport and cycling: Reproducible methods, results and actionable evidence from biclaR

### 2024-05-07 paper/PaperCEUS/PaperCEUS_rev1.Rmd
respond to 2.2, 2.3, 2.4, 2.5

Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can replace a high proportion of private car trips [@MARTENS2007326].
In The Netherlands, which has the highest mode share of cycling in the world, cycling accounts for more than a third of all trips to and from rail stations at the 'home' end of the journey, greatly increasing the ability of the transport system [@RIETVELD200071].
This approach to reducing car dependency and associated externalities requires interventions and programs to make bicycling more appealing [@lapaix_role_2021].
The resulting public investments can have significant social and environmental benefits [@internationaltransportforum_integrating_2017].
Despite the benefits of cycling-PT intermodality, the potential of this combination is often overlooked in transport planning [@lapaix_role_2021].
<!-- Many studies refer to the potential of cycling and bike-sharing to replace car and PT trips, but not to the potential of serving as first and last miles. -->

@@

@@

@@

This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA, while achieving the national cycling targets.
After presenting the methods used, it assesses its socio-environmental impacts using open data and open-source tools.

@@

@@

The intermodality scenario considers trips made by PT in which cycling is used for the first and last legs.
We restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination) or up to 25 minutes on bike travel-time.
This conservative approach was adopted to capture the fact that cycling stages as part of a multi-modal trip are likely to be shorter than cycling-only trips [@vanmil_insights_2021].
Furthermore, we have imposed restrictions on PT usage, limiting it to trips without PT transfers, and within a duration of up to 2 hours (120 minutes).
Additionally, we have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
These restrictions can be eased in the future when testing more developed policy interventions to enhance intermodality between cycling and PT, considering both the vehicle and infrastructure perspectives.

@@

```{r mapzones, fig.cap="Potential Bike + PT daily trips for the 10% target, by municipality"}
CAR_PT_TRANSF = readRDS("load/CAR_TP_TRANSF.Rds")

@@

Our findings show that cycling *in combination* with PT could replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution, based on conservative assumptions.
Although this paper does not address the designing specific solutions, our findings indicate a potential surge in demand (+6.3%) of public transport interfaces to align with the national strategies.
This added demand is conceptually driven from individuals utilising bicycles, either personal or shared, which would be parked at the PT stations.
Consequently, operators and public space managers would face the challenge of accommodating this increased demand for bicycle parking and/or docks from a bike sharing system.
Additionally, they may need to explore options to enhance bicycle carriage capacity on public transport services, ensuring alignment with national targets.

## Model replicability and complexity

This analysis can be applied to other metropolitan areas, providing a basis for decision-making and policy development to support a modal shift from car to PT. This process involves a minimum of data requirements, such as the origin-destination data, road network, and the PT GTFS data, which are available in most metropolitan areas.
Furthermore, it is essential to incorporate existing cycling level targets specific to those areas or consider the possibility of utilising the same targets as outlined in the Portuguese national strategy, in order to compare results with different areas.
Nevertheless, the extent of the potential shift from car to PT is highly dependent on the distances typically travelled in such contexts, and the availability of PT services - which can vary significantly between regions, and for which we are not assuming any change in this study.

<!-- CONTINUE HERE -->
A further refinement could be to differentiate between ‘simpler’ no-transfer PT trips (which require less person effort) and more arduous multi-transfer PT trips, using the same methods: we could set a different maximum cycling stage distance for the former and a shorter maximum cycling stage distance for the latter. However, this would have an impact on the explainability of the results to policy makers and the overall complexity.
We now discuss this and other potential refinements to the methods using the same input datasets and open, reproducible methods in the discussion with reference to the **likelihood of diminishing returns with additional model complexity**.

@@

>>>>

Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can replace a high proportion of private car trips [@MARTENS2007326; @van2021insights].
In The Netherlands, which has the highest mode share of cycling in the world, cycling accounts for more than a third of all trips to and from rail stations at the 'home' end of the journey, greatly increasing the ability of the transport system [@RIETVELD200071].
This approach to reducing car dependency and associated externalities requires interventions and programs to make bicycling more appealing [@lapaix_role_2021].
The resulting public investments can have significant social and environmental benefits [@internationaltransportforum_integrating_2017].
Despite the benefits of cycling-PT intermodality, the potential of this combination is often overlooked in transport planning [@lapaix_role_2021].
<!-- Many studies refer to the potential of cycling and bike-sharing to replace car and PT trips, but not to the potential of serving as first and last miles. -->

@@

Regarding cycling trips in the LMA, 55% are up to 5 km, and 88% are up to 10 km.
These values are not in line with the typically found in cycling distance decay curves, which show a high percentage of trips are up to 5 km (~75%), and a smaller proportion up to 10 km [@krizek2007detailed; @larsen2010beyond; @DDfunction2023].
This may be due to the low cycling modal share in the LMA - even more limited in the sample survey, which may be biased towards shorter trips.

@@

**TO-DO**

-   mais mapas:
-   Densidade populational - nº de viagens
-   Pares OD - desire lines width

@@

This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA.
After presenting the methods used, we assess its socio-environmental impacts using open data and open-source tools.

It is essential to note that our approach is not predictive but scenario-based, providing a range of potential outcomes regarding the target (% of cycling tips) and the routing profile (safe or direct routes), rather than a single forecast.
These scenarios are rooted in the explicit national cycling targets, assuming that the potential for cycling is proportional to the reduction of car trips.
Thus said, we do not predict the generation of new trips, but the substitution of car trips by cycling in combination with PT, considering the potential of the first and last mile of these journeys to be made by bicycle.
Additionally, we do not model changes in the baseline PT trips, under the implicit assumption that the initial and final stages of these trips remain unaltered.
A strength of this _backcasting_ approach is that we can simulate the aggregate travel patterns that would meet particular policy objectives, and explore what the implications could be for the public transport system and provision of safe cycling infrastructure to public transport nodes at regional, local authority and, importantly from a local transport planning practitioner perspective, corridor level.

@@

## Case Study

@@

The intermodality scenario considers trips made by PT in which cycling is used for the first and last legs.
We restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination).
Furthermore, we have imposed restrictions on PT usage, limiting it to trips without PT transfers, and within a duration of up to 2 hours (120 minutes).
Additionally, we have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).

This conservative approach was adopted to capture the fact that cycling stages as part of a multi-modal trip are likely to be shorter than cycling-only trips [@vanmil_insights_2021].
Additionally, @Leferink2017 found that the bicycle is most attractive option as a first and last mile mode when the distance ranges from 3 to 5 km, as observed across various comparative studies.
This restrictions are based on the assumption that 5 km is a plausible maximum distance that people are willing to travel by bike as part of a longer journey.
This applies especially for cycling beginners shifting from cars, who are less likely to be confident and experienced cyclists than the existing cyclists.

These restrictions can be eased in the future when testing more developed policy interventions to enhance intermodality between cycling and PT, considering both the vehicle and infrastructure perspectives.

@@

```{r mapzones, eval=FALSE, fig.cap="Potential Bike + PT daily trips for the 10% target, by municipality"}
CAR_PT_TRANSF = readRDS("load/CAR_TP_TRANSF.Rds")

@@

Our findings show that cycling *in combination* with PT could replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution, based on conservative assumptions.
Although this paper does not address the designing specific solutions, our findings indicate a potential surge in demand (+6.3%) of public transport interfaces to align with the national strategies.
This added demand is conceptually driven from individuals utilising bicycles, either personal or shared, which would be parked at the PT stations.
Consequently, operators and public space managers would face the challenge of accommodating this increased demand for bicycle parking and/or docks from a bike sharing system, since this availability plays a crucial role in the decision-making process of individuals to use combined bike-train [@jonkeren2021bicycle].
Additionally, they may need to explore options to enhance bicycle carriage capacity on public transport services, ensuring alignment with national targets.

## Model replicability and limitations

@@

This analysis can be applied to other metropolitan areas, providing a basis for decision-making and policy development to support a modal shift from car to PT.
This process involves a minimum of data requirements, such as the origin-destination data, road network, and the PT GTFS data, which are available in most metropolitan areas.
Furthermore, it is essential to incorporate existing cycling level targets specific to those areas or consider the possibility of utilising the same targets as outlined in the Portuguese national strategy, in order to compare results with different areas.

Nevertheless, the extent of the potential shift from car to PT is highly dependent on the distances typically travelled in such contexts, and the availability of PT services - which can vary significantly between regions, and for which we are not assuming any change in this study.
For instance, we considered only the inter-municipal bus services that accommodate bicycles on board.
Future research could explore the potential of equipping other regional bus services with this feature and analyse their attractiveness, in conjunction with cycling, to address current car trip demands.

A strength of the approach is that there is no requirement for sensitive individual trip level data to reproduce the approach in other cities.
However, we acknowledge that this approach has limitations in terms of being able to identify exactly which trips are being replaced -- in terms of the age of people making the trips, the types of car they drive, and other individual or vehicle level variables.
We assumed uniformity in the type of vehicle for all replaced car journeys, albeit with variations in the fuels utilized.
Additionally, we assumed consistency in the health impacts across all individuals for the replaced car trips
This approach is a simplification of the real-world scenario, but it is a necessary one to make the analysis feasible and reproducible.

A further refinement could be to differentiate between ‘simpler’ no-transfer PT trips (which require less person effort) and more arduous multi-transfer PT trips, using the same methods, by relaxing the assumption of no-transfers between PT modes in `r5r`.
This would led to a larger number of total daily trips that can be made by cycling + PT, in particular in locations where the PT network is less complex and requires more transfers.
We could also set a different maximum lenght of cycling stage, extending the bicycle catchment area.
However, despite the added value of a sensibility analyses, they would also increase the overall complexity, and would have an impact on the explainability of the results to be used by policy makers.

### 2024-05-07 paper/PaperCEUS/PaperCEUS_rev1.tex
respond to 2.2, 2.3, 2.4, 2.5

  \title{Modeling the impacts of replacing car trips with combined
public transport and cycling: Reproducible methods, results and
actionable evidence from biclaR}
    \author[CERIS]{Rosa Félix\corref{cor1}%
  \corref{cor1}%
  }
   \ead{rosamfelix@tecnico.ulisboa.pt} 
    \author[CERIS]{Filipe Moura%
  %
  }
   \ead{fmoura@tecnico.ulisboa.pt} 
    \author[ITS]{Robin Lovelace%
  %
  }
   \ead{r.lovelace@uleeds.uk} 
      \affiliation[CERIS]{CERIS, Instituto Superior Técnico -
Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
    \affiliation[ITS]{Institute for Transport Studies, University of
Leeds. 34-40 University Rd, Leeds LS2 9JT, UK}
    \cortext[cor1]{Corresponding author}

@@

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can replace a high proportion of private car
trips \citep{MARTENS2007326}. In The Netherlands, which has the highest
mode share of cycling in the world, cycling accounts for more than a
third of all trips to and from rail stations at the `home' end of the
journey, greatly increasing the ability of the transport system
\citep{RIETVELD200071}. This approach to reducing car dependency and
associated externalities requires interventions and programs to make
bicycling more appealing \citep{lapaix_role_2021}. The resulting public
investments can have significant social and environmental benefits
\citep{internationaltransportforum_integrating_2017}. Despite the
benefits of cycling-PT intermodality, the potential of this combination
is often overlooked in transport planning \citep{lapaix_role_2021}.

The potential of cycling as a complementary mode of PT is substantial
worldwide, especially in cities with established public transport
networks or substantial ambitions to develop them. In the Lisbon
metropolitan area (LMA) the largest metropolitan area in Portugal, the
modal share of cycling is low, but the potential for cycling as a
complementary mode of PT is high. According to the latest mobility
survey conducted in 2018 \citep{IMOB}, the LMA registered a total of 5.3
million daily trips, with only 0.5\% by bicycle. Car modal share was
58.4\%, while PT accounted for 15.5\% (see Figure \ref{fig:mododist}).
The number of intra-municipal trips --- with origin and destination in
the same municipality --- amounts to 3.5 million trips. This exceeds the
number of inter-municipal trips (1.8 million trips), involving travel
between different municipalities. Cars and public transport are the most
used modes for intercity trips, with cars being the predominant choice
for all journeys.\\
53\% trips are up to 5 km distance, and 71\% up to 10 km. Nevertheless,
29\% of trips are longer than 10 km, which requires the use of motorized
modes, or active modes in combination with public transport.

@@

@@

This paper estimates the potential for combining cycling and PT to
substitute car trips in the LMA, while achieving the national cycling
targets. After presenting the methods used, it assesses its
socio-environmental impacts using open data and open-source tools.

@@

@@

The intermodality scenario considers trips made by PT in which cycling
is used for the first and last legs. We restricted our analysis to the
first and last legs with a combined length of up to 5 km (for example: 1
km from origin to interface \emph{A} plus 4 km from interface \emph{B}
to destination) or up to 25 minutes on bike travel-time. This
conservative approach was adopted to capture the fact that cycling
stages as part of a multi-modal trip are likely to be shorter than
cycling-only trips \citep{vanmil_insights_2021}. Furthermore, we have
imposed restrictions on PT usage, limiting it to trips without PT
transfers, and within a duration of up to 2 hours (120 minutes).
Additionally, we have only included PT modes that can easily accommodate
bicycles, such as trains, ferries, trams, and inter-municipal bus lines
equipped with bike racks (Figure \ref{fig:map1}). These restrictions can
be eased in the future when testing more developed policy interventions
to enhance intermodality between cycling and PT, considering both the
vehicle and infrastructure perspectives.

@@

Our findings show that cycling \emph{in combination} with PT could
replace 10\% of current LMA trips, with an additional 6\% of PT journeys
prone to further substitution, based on conservative assumptions.
Although this paper does not address the designing specific solutions,
our findings indicate a potential surge in demand (+6.3\%) of public
transport interfaces to align with the national strategies. This added
demand is conceptually driven from individuals utilising bicycles,
either personal or shared, which would be parked at the PT stations.
Consequently, operators and public space managers would face the
challenge of accommodating this increased demand for bicycle parking
and/or docks from a bike sharing system. Additionally, they may need to
explore options to enhance bicycle carriage capacity on public transport
services, ensuring alignment with national targets.

This analysis can be applied to other metropolitan areas, providing a
basis for decision-making and policy development to support a modal
shift from car to PT. This process involves a minimum of data
requirements, such as the origin-destination data, road network, and the
PT GTFS data, which are available in most metropolitan areas.
Furthermore, it is essential to incorporate existing cycling level
targets specific to those areas or consider the possibility of utilising
the same targets as outlined in the Portuguese national strategy, in
order to compare results with different areas. Nevertheless, the extent
of the potential shift from car to PT is highly dependent on the
distances typically travelled in such contexts, and the availability of
PT services - which can vary significantly between regions, and for
which we are not assuming any change in this study. This paper focus
primarily on the potential of making cycling and PT a viable alternative
to car trips, and the socio-environmental benefits of such a shift.

>>>>

  \title{Modeling the impacts of replacing car trips with combined
public transport and cycling: Reproducible methods, results and
actionable evidence from biclaR}
    \author[CERIS]{Rosa Félix\corref{cor1}%
  \corref{cor1}%
  }
   \ead{rosamfelix@tecnico.ulisboa.pt} 
    \author[CERIS]{Filipe Moura%
  %
  }
   \ead{fmoura@tecnico.ulisboa.pt} 
    \author[ITS]{Robin Lovelace%
  %
  }
   \ead{r.lovelace@uleeds.uk} 
      \affiliation[CERIS]{CERIS, Instituto Superior Técnico -
Universidade de Lisboa. Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
    \affiliation[ITS]{Institute for Transport Studies, University of
Leeds. 34-40 University Rd, Leeds LS2 9JT, UK}
    \cortext[cor1]{Corresponding author}

@@

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can replace a high proportion of private car
trips \citep{MARTENS2007326, van2021insights}. In The Netherlands, which
has the highest mode share of cycling in the world, cycling accounts for
more than a third of all trips to and from rail stations at the `home'
end of the journey, greatly increasing the ability of the transport
system \citep{RIETVELD200071}. This approach to reducing car dependency
and associated externalities requires interventions and programs to make
bicycling more appealing \citep{lapaix_role_2021}. The resulting public
investments can have significant social and environmental benefits
\citep{internationaltransportforum_integrating_2017}. Despite the
benefits of cycling-PT intermodality, the potential of this combination
is often overlooked in transport planning \citep{lapaix_role_2021}.

The potential of cycling as a complementary mode of PT is substantial
worldwide, especially in cities with established public transport
networks or substantial ambitions to develop them. In the Lisbon
metropolitan area (LMA) the largest metropolitan area in Portugal, the
modal share of cycling is low, but the potential for cycling as a
complementary mode of PT is high. According to the latest mobility
survey conducted in 2018 \citep{IMOB}, the LMA registered a total of 5.3
million daily trips, with only 0.5\% by bicycle. Car modal share was
58.4\%, while PT accounted for 15.5\% (see Figure \ref{fig:mododist}).
The number of intra-municipal trips --- with origin and destination in
the same municipality --- amounts to 3.5 million trips. This exceeds the
number of inter-municipal trips (1.8 million trips), involving travel
between different municipalities. Cars and public transport are the most
used modes for intercity trips, with cars being the predominant choice
for all journeys. 53\% trips are up to 5 km distance, and 71\% up to 10
km. Nevertheless, 29\% of trips are longer than 10 km, which requires
the use of motorized modes, or active modes in combination with public
transport.

Regarding cycling trips in the LMA, 55\% are up to 5 km, and 88\% are up
to 10 km. These values are not in line with the typically found in
cycling distance decay curves, which show a high percentage of trips are
up to 5 km (\textasciitilde75\%), and a smaller proportion up to 10 km
\citep{krizek2007detailed, larsen2010beyond, DDfunction2023}. This may
be due to the low cycling modal share in the LMA - even more limited in
the sample survey, which may be biased towards shorter trips.

@@

\textbf{TO-DO}

\begin{itemize}
\tightlist
\item
  mais mapas:
\item
  Densidade populational - nº de viagens
\item
  Pares OD - desire lines width
\end{itemize}

@@

This paper estimates the potential for combining cycling and PT to
substitute car trips in the LMA. After presenting the methods used, we
assess its socio-environmental impacts using open data and open-source
tools.

It is essential to note that our approach is not predictive but
scenario-based, providing a range of potential outcomes regarding the
target (\% of cycling tips) and the routing profile (safe or direct
routes), rather than a single forecast. These scenarios are rooted in
the explicit national cycling targets, assuming that the potential for
cycling is proportional to the reduction of car trips. Thus said, we do
not predict the generation of new trips, but the substitution of car
trips by cycling in combination with PT, considering the potential of
the first and last mile of these journeys to be made by bicycle.
Additionally, we do not model changes in the baseline PT trips, under
the implicit assumption that the initial and final stages of these trips
remain unaltered. A strength of this \emph{backcasting} approach is that
we can simulate the aggregate travel patterns that would meet particular
policy objectives, and explore what the implications could be for the
public transport system and provision of safe cycling infrastructure to
public transport nodes at regional, local authority and, importantly
from a local transport planning practitioner perspective, corridor
level.

@@

\subsection{Case Study}\label{case-study}

@@

The intermodality scenario considers trips made by PT in which cycling
is used for the first and last legs. We restricted our analysis to the
first and last legs with a combined length of up to 5 km (for example: 1
km from origin to interface \emph{A} plus 4 km from interface \emph{B}
to destination). Furthermore, we have imposed restrictions on PT usage,
limiting it to trips without PT transfers, and within a duration of up
to 2 hours (120 minutes). Additionally, we have only included PT modes
that can easily accommodate bicycles, such as trains, ferries, trams,
and inter-municipal bus lines equipped with bike racks (Figure
\ref{fig:map1}).

This conservative approach was adopted to capture the fact that cycling
stages as part of a multi-modal trip are likely to be shorter than
cycling-only trips \citep{vanmil_insights_2021}. Additionally,
\citet{Leferink2017} found that the bicycle is most attractive option as
a first and last mile mode when the distance ranges from 3 to 5 km, as
observed across various comparative studies. This restrictions are based
on the assumption that 5 km is a plausible maximum distance that people
are willing to travel by bike as part of a longer journey. This applies
especially for cycling beginners shifting from cars, who are less likely
to be confident and experienced cyclists than the existing cyclists.

These restrictions can be eased in the future when testing more
developed policy interventions to enhance intermodality between cycling
and PT, considering both the vehicle and infrastructure perspectives.

@@

Our findings show that cycling \emph{in combination} with PT could
replace 10\% of current LMA trips, with an additional 6\% of PT journeys
prone to further substitution, based on conservative assumptions.
Although this paper does not address the designing specific solutions,
our findings indicate a potential surge in demand (+6.3\%) of public
transport interfaces to align with the national strategies. This added
demand is conceptually driven from individuals utilising bicycles,
either personal or shared, which would be parked at the PT stations.
Consequently, operators and public space managers would face the
challenge of accommodating this increased demand for bicycle parking
and/or docks from a bike sharing system, since this availability plays a
crucial role in the decision-making process of individuals to use
combined bike-train \citep{jonkeren2021bicycle}. Additionally, they may
need to explore options to enhance bicycle carriage capacity on public
transport services, ensuring alignment with national targets.

\subsection{Model replicability and
limitations}\label{model-replicability-and-limitations}

This paper focus primarily on the potential of making cycling and PT a
viable alternative to car trips, and the socio-environmental benefits of
such a shift.

This analysis can be applied to other metropolitan areas, providing a
basis for decision-making and policy development to support a modal
shift from car to PT. This process involves a minimum of data
requirements, such as the origin-destination data, road network, and the
PT GTFS data, which are available in most metropolitan areas.
Furthermore, it is essential to incorporate existing cycling level
targets specific to those areas or consider the possibility of utilising
the same targets as outlined in the Portuguese national strategy, in
order to compare results with different areas.

Nevertheless, the extent of the potential shift from car to PT is highly
dependent on the distances typically travelled in such contexts, and the
availability of PT services - which can vary significantly between
regions, and for which we are not assuming any change in this study. For
instance, we considered only the inter-municipal bus services that
accommodate bicycles on board. Future research could explore the
potential of equipping other regional bus services with this feature and
analyse their attractiveness, in conjunction with cycling, to address
current car trip demands.

A strength of the approach is that there is no requirement for sensitive
individual trip level data to reproduce the approach in other cities.
However, we acknowledge that this approach has limitations in terms of
being able to identify exactly which trips are being replaced -- in
terms of the age of people making the trips, the types of car they
drive, and other individual or vehicle level variables. We assumed
uniformity in the type of vehicle for all replaced car journeys, albeit
with variations in the fuels utilized. Additionally, we assumed
consistency in the health impacts across all individuals for the
replaced car trips This approach is a simplification of the real-world
scenario, but it is a necessary one to make the analysis feasible and
reproducible.

A further refinement could be to differentiate between `simpler'
no-transfer PT trips (which require less person effort) and more arduous
multi-transfer PT trips, using the same methods, by relaxing the
assumption of no-transfers between PT modes in \texttt{r5r}. This would
led to a larger number of total daily trips that can be made by cycling
+ PT, in particular in locations where the PT network is less complex
and requires more transfers. We could also set a different maximum
lenght of cycling stage, extending the bicycle catchment area. However,
despite the added value of a sensibility analyses, they would also
increase the overall complexity, and would have an impact on the
explainability of the results to be used by policy makers.
